The all-seeing eye of Auger electron spectroscopy: a study on aqueous KCl

### 2018-02-05 abstract.tex
jan 24 (denis' corrections & tsveta's)
Shift explained as resulting from internal ionization, an argument for
the attribution of island C as 2p-2 3d state based on the CT paper
feb 5 (nico's corrections)
Nico's final corrections included; new graphical abstract

%Kolya and Nico corrected Jan 8
X-ray absorption and Auger electron spectroscopies are powerful tools to probe the electronic structure and immediate surroundings of ions in a solution. In this work we use a combination of these spectroscopies to study the electronic structure and decay in aqueous KCl at the K-edges of \ki~and \cli. Although the two ions are isoelectronic, their Auger electron spectra as a function of the photon energy exhibit notably different features. To explain these differences, we carried out {\it ab initio} calculations of both the core excited states and the final Auger states of \ki, \cli~and their microsolvated clusters. Our calculations show that the energy order of the 3d and 4p orbitals is inverted in \ki~with respect to \cli. The reverse orbital order in the two ions is reflected in both the ordering of the core excited states and in the final states populated in the resonant Auger decay. The energetic proximity of the 3d and 4p virtual states in the bare \ki~ion leads to their mixing in the presence of the solvent, and to the population of the dipole forbidden 1s$\,\rightarrow\,$3d state upon K-shell excitation in an aqueous solution. The resonant Auger decay of this state results in a separate feature in the Auger electron spectrum of \ki~which is absent in the spectrum of \cli.

>>>>

%Kolya and Nico corrected Jan 8
X-ray absorption and Auger electron spectroscopies are powerful tools to probe the electronic structure and immediate surroundings of ions in a solution. In this work we use a combination of these methods to study the electronic structure and decay in aqueous KCl at the K-edges of \ki~and \cli. Although the two ions are isoelectronic, their Auger electron spectra as a function of the photon energy exhibit notably different features. To explain these differences, we carried out {\it ab initio} calculations of both the core excited states and the final Auger states of \ki, \cli~and their microsolvated clusters. Our calculations show that the energy order of the 3d and 4p orbitals is inverted in \ki~with respect to \cli. The reverse orbital order in the two ions is reflected in both the ordering of the core excited states and in the final states populated in the resonant Auger decay. The energetic proximity of the 3d and 4p virtual states in the bare \ki~ion leads to their mixing in the presence of the solvent, and to the population of the dipole forbidden 1s$\,\rightarrow\,$3d state upon K-shell excitation in an aqueous solution. The resonant Auger decay of this state results in a separate feature in the Auger electron spectrum of \ki~which is absent in the spectrum of \cli.
{\color{red}The results of this work represent a pioneering study of the decay processes initiated by photoabsorption in the tender x-ray regime close to threshold in liquids and are thus of importance in unveiling the mechanisms of radiation damage in biologically relevant systems.}
%Denis: line 2, I replaced spectroscopies by methods

### 2018-02-05 conclusion.tex
feb 5 (nico's corrections)
Nico's final corrections included; new graphical abstract

Using a combination of x-ray absorption and Auger electron spectroscopy in the tender x-ray regime, in this work we studied the electronic structure of aqueous solution of KCl at the K-edges of both K and Cl. The Auger electron spectra of both ions as a function of photon energy exhibit features of normal as well as resonant Auger processes. To interpret the resonant Auger features in the experimental spectrum, we performed {\it ab initio} calculations on microsolvated clusters of \ki~and \cli. Our calculations show that the energy ordering of the 3d and 4p virtual orbitals of \cli~is inverted compared to \ki, and also that the energy splitting between the bright 1s$\,\rightarrow\,$4p and dark 1s$\,\rightarrow\,$3d core excited states is larger in the chlorine case. Thus, the energetic proximity of the 3d and 4p orbitals in the bare \ki~ion results in the dipole forbidden 1s$\,\rightarrow\,$3d state acquiring intensity in a solution as a result of mixing with the dipole allowed 1s$\,\rightarrow\,$4p excitation. The spectator Auger decay of this state produces an additional dispersive feature which is manifest as a separate island in the Auger electron spectrum at high kinetic energies. In the case of \cli~the two core excited states do not interact, and therefore, only fingerprints of the population and Auger decay of the dipole allowed 1s$\,\rightarrow\,$4p state are observed in the spectrum. Moreover, using the core-hole clock method we estimated the time of delocalization of the core excited electron at the pre-edge region of both ions. Whereas in \ki$_{\text{aq}}$ the delocalization of the core excited electron is slower than the resonant Auger process, in the case of \cli$_{\text{aq}}$ the two processes occur on a comparable timescale.

>>>>

Using a combination of x-ray absorption and Auger electron spectroscopy in the tender x-ray regime, in this work we studied the electronic structure of aqueous solution of KCl at the K-edges of both K and Cl. The Auger electron spectra of both ions as a function of photon energy exhibit features of normal as well as resonant Auger processes. To interpret the resonant Auger features in the experimental spectrum, we performed {\it ab initio} calculations on microsolvated clusters of \ki~and \cli. Our calculations show that the energy ordering of the 3d and 4p virtual orbitals of \cli~is inverted compared to \ki, and also that the energy splitting between the bright 1s$\,\rightarrow\,$4p and dark 1s$\,\rightarrow\,$3d core excited states is larger in the chlorine case. Thus, the energetic proximity of the 3d and 4p orbitals in the bare \ki~ion results in the dipole forbidden 1s$\,\rightarrow\,$3d state acquiring intensity in a solution as a result of mixing with the dipole allowed 1s$\,\rightarrow\,$4p excitation. The spectator Auger decay of this state produces an additional dispersive feature which is manifest as a separate island in the Auger electron spectrum at high kinetic energies. In the case of \cli~the two core excited states do not interact, and therefore, only fingerprints of the population and Auger decay of the dipole allowed 1s$\,\rightarrow\,$4p state are observed in the spectrum. {\color{red}Moreover, using the core-hole clock method we estimated the time of delocalization of the core excited electron at the pre-edge region of \cli$_{\text{aq}}$. Our results show that in this case, the resonant Auger decay and the delocalization of the excited electron occur on a comparable timescale. In the case of \ki$_{\text{aq}}$, we cannot make an accurate estimate, however, one can expect a much less efficient delocalization of the core excited electron.}

### 2018-02-05 introduction.tex
feb 2 (version sent to co-authors)
Changes made by Denis, new figures for SI, text changed.
feb 5 (nico's corrections)
Nico's final corrections included; new graphical abstract

X-ray absorption and Auger electron spectroscopies are powerful tools to study the electronic structure and the nearest environment of atoms and molecules in gas, liquid and solid phase. Understanding how atoms or molecules respond to irradiation with x-rays gives insight into the structure of solutions (Ref.\ \citep{smith17:13909} and references therein), and the mechanism of radiation damage \citep{ONeill02:329,Carugo05:213,Stumpf16:237}. Upon absorption of an x-ray photon, core excited or core ionized states of a specific atom are populated depending on the photon energy. The relaxation of these highly energetic states involves an ultrafast cascade of intraatomic processes, such as radiative and Auger decays. Furthermore, if the initially excited or ionized species is embedded in an environment, interatomic processes such as charge and energy transfer \citep{Pokapanich09:7264,Pokapanich11:13430,Stumpf16:237,unger17:708,ceolin17:263003} are possible.

The course of a decay cascade depends on the character of the initially populated states. This has been well understood in atoms and molecules in gas phase  \citep{stoychev08:074307,Demekhin08:043421,Demekhin09:104303,Ouchi11:053415,Miteva14:164303,travnikova16:213001}. In the case of a core ionized state, the Auger decay process, designated as normal Auger decay (see Fig.\ \ref{fg:auger}), leads to the population of doubly ionized final states localized on the initially ionized unit {\color{red}\citep{stoychev08:074307,Demekhin08:043421,Demekhin09:104303,Ouchi11:053415}}. Auger processes in rare gas clusters have been also investigated (refs). The normal Auger decay process in clusters proceeds similarly to that in atoms or molecules. However, in the case of a core excited state, the resonant Auger process competes with the process of delocalization of the excited electron in clusters. If the initially core excited electron delocalizes within the lifetime of the core hole, then normal instead of resonant Auger decay is observed \citep{Bjorneholm95:3017}.

@@

\begin{figure}
\includegraphics{figures/auger_process.eps}
\caption{Schematic representation of the normal and resonant Auger processes of the isoelectronic \ki~and \cli~ions. The charges of the final Auger states in the two ions are also shown.}
\label{fg:auger}
\end{figure}

>>>>

X-ray absorption and Auger electron spectroscopies are powerful tools to study the electronic structure and the nearest environment of atoms and molecules in gas, liquid and solid phase. Understanding how atoms or molecules respond to irradiation with x-rays gives insight into the structure of solutions (see Ref.\ \citep{smith17:13909} and references therein), and the mechanism of radiation damage \citep{ONeill02:329,Carugo05:213,Stumpf16:237}. Upon absorption of an x-ray photon, core excited or core ionized states of a specific atom are populated depending on the photon energy. The relaxation of these highly energetic states involves an ultrafast cascade of intraatomic processes, such as radiative and Auger decays. Furthermore, if the initially excited or ionized species is embedded in an environment, interatomic processes such as charge and energy transfer \citep{Pokapanich09:7264,Pokapanich11:13430,Stumpf16:237,unger17:708,ceolin17:263003} are possible.

The course of a decay cascade depends on the character of the initially populated states. This has been well understood in atoms and molecules in gas phase  \citep{stoychev08:074307,Demekhin08:043421,Demekhin09:104303,Ouchi11:053415,Miteva14:164303,travnikova16:213001}. In the case of a core ionized state, the Auger decay process, designated as normal Auger decay (see Fig.\ \ref{fg:auger}), leads to the population of doubly ionized final states localized on the initially ionized unit \citep{stoychev08:074307,Demekhin08:043421,Demekhin09:104303,Ouchi11:053415}. Auger processes in rare gas clusters have been also investigated (refs). The normal Auger decay process in clusters proceeds similarly to that in atoms or molecules. However, in the case of a core excited state, the resonant Auger process competes with the process of delocalization of the excited electron in clusters. If the initially core excited electron delocalizes within the lifetime of the core hole, then normal instead of resonant Auger decay is observed \citep{Bjorneholm95:3017}.

@@

\begin{figure}
\includegraphics[scale=0.8]{figures/auger_process.pdf}
\caption{Schematic representation of the normal and resonant Auger processes of the isoelectronic \ki~and \cli~ions. The charges of the final Auger states in the two ions are also shown.}
\label{fg:auger}
\end{figure}

### 2018-02-05 methods.tex
feb 5 (nico's corrections)
Nico's final corrections included; new graphical abstract

For the present experiment we used the newly operational microjet setup that was specifically designed for the HAXPES station of the GALAXIES beamline \citep{ceolin13:188,rueff15:175}. A differentially-pumped tube in which the microjet head is inserted, is mounted on a 3-axes manipulator in front of the spectrometer lens. Two holes of 2\,mm diameter allow the photons to go in and out. At the end of the tube and in front of the lens, a 500\,$\mu$m diameter hole skimmer allows the electrons created at the interaction point to go in the direction of the spectrometer. The microjet head is mostly composed of a 30\,$\mu$m diameter vertical glass capillary facing a temperature-controlled catcher in CuBe having a 300\,$\mu$m hole, and a camera. Piezo motors allow their precise alignment relative to each other and to the photon beam. The catcher is placed at a distance of about 5\,mm from the capillary and is permanently pumped in order to extract the liquid. For the present experiment, a 0.5M KCl aqueous solution is injected in the capillary by a HPLC pump with a constant flux of 1.6\,ml/min. The alignment of the setup is performed on a salted aqueous solution by measuring the O1s XPS peak intensity and by optimizing the liquid phase vs gas phase ratio. The pressure in the main chamber is kept below the 10$^{-5}$\,mbar range whereas it is kept at about 10$^{-4}$\,mbar in the differentially-pumped tube when the HPLC pump is ON. Our equipment is an updated version of the equipment used in the reference \cite{faubel88:269}. The aqueous potassium chloride solution was prepared by mixing >99\% KCl salt with deionized water. Filtering and degazing procedures were systematically performed before injecting the solution. The spectrometer resolution of about 0.6\,eV was achieved with the 500\,eV pass energy and 0.5\,mm slits. The photon energy resolution achieved at 2.8\,keV and 3.6\,keV was about 0.3\,eV and 0.4\,eV, respectively

@@

The theoretical X-ray absorption spectra were computed for the hexa-coordinated clusters of both ions, \ki(H$_2$O)$_6$ and \cli(H$_2$O)$_6$, which can be considered as representatives of the complete first solvation shell of the two ions \citep{Ohtaki93:1157,soper06:180,ma14:1006}. The two structures were optimized at the DFT level of theory using the B3LYP functional and the 6-311++G(2d,2p) basis set \citep{Krishnan80:650,Blaudeau97:5016}. The geometry optimization was performed with the Gaussian 09 package \citep{g09}. In order to obtain more realistic structures corresponding to the bulk solution, we carried out constrained geometry optimization starting with the equilibrium gas-phase geometries \citep{lee99:3995}(REF for Cl-) belonging to the D$_3$ point group and increasing the angle $\theta$ between the K-O/Cl-O bond and the $C_3$ axis to 55$^{\circ}$ in the case of \ki~and 50$^{\circ}$ in the case of \cli. This angle was chosen such that the O-K-O and O-Cl-O angles are around the maxima of the angular distributions obtained from quantum mechanics / molecular mechanics dynamical simulations in Ref.\ \citep{ma14:1006}. Moreover, we fixed the K-O and Cl-O distances to 2.840~\AA~and 3.140~\AA, respectively, such that they correspond to the distances from other theoretical and experimental works \citep{ge13:13169,gora00:7,Ohtaki93:1157,soper06:180,ma14:1006}. The hexa-coordinated structures are presented in Fig.\ \ref{fg:xas_kcl}. %In both cases, they belong D$_3$ point group \citep{rao08:12944,ge13:13169}.

The energies and transition moments of the core excited states of the microsolvated clusters were computed with the algebraic diagrammatic construction method for the polarization propagator \citep{sch82:2395} within the core-valence separation approximation \citep{bar85:867,ced80:206,ced81:1038} (CVS-ADC(2)x) as implemented in the Q-Chem package \citep{Wenzel14:1900,Wenzel14:4583,Wormit14:774,QChem2015}. In the case of \cli~the 6-311++G(3df,3pd) basis set \citep{Krishnan80:650,McLean80:5639} (excluding the f functions) was used for all atoms, whereas in the case of \ki~we used the 6-311+G(2d,p) basis set \citep{Krishnan80:650,Blaudeau97:5016} on all atoms, and two additional sets of s, p and d diffuse functions were added on K. In our calculations the core space comprises the 1s orbital of K$^{+}$ or \cli, whereas the remaining occupied orbitals are included in the valence space. For the calculations of the XAS spectra we used the C$_2$ point group in the case of K$^{+}$(H$_2$O)$_6$ and \cli(H$_2$O)$_6$. To account for the lifetime broadening due to the Auger decay of the core excited states, we convolved the theoretical spectra with a Lorentzian of FWHM 0.74\,eV and 0.62\,eV in the case of \ki~and \cli, respectively \citep{Krause79:329}. We analyzed the core excited states by expanding the singly occupied natural orbitals (SONOs) $\psi_{i}$ of the microsolvated clusters in the basis of SONOs of the bare K$^{+}$ or \cli~ion $\chi_{nl}$
%
\begin{equation}\label{eq:sono_proj}
\psi_{i} = \sum_{nl} a^{i}_{nl} \chi_{nl}
\end{equation}
%
where $n$ and $l$ stand for the principal and orbital quantum numbers as described in Ref.\ \citep{miteva16:16671}. The expansion coefficients $a^{i}_{nl}$ show the degree of delocalization of the excited electron and the mixing of the core excited states in the crystal field created by the surrounding water molecules (see Fig.\ \ref{fg:xas_kcl}).

The final states following KLL resonant Auger decay of K$^{+}$(H$_2$O)$_6$ and \cli(H$_2$O)$_6$ were computed at the Configuration Interaction Singles (CIS) level using the Graphical Unitary Group Approach (GUGA) as implemented in the GAMESS-US package \citep{GUGA_PhysScr_21,GUGA_JCP_70,GUS}. In order to account for the relaxation effects upon core ionization, we used a restricted open-shell Hartree-Fock reference wave function with a hole in the 2s orbital of both \ki~and \cli.  We used the 6-311++G(2d,2p) basis set \citep{Krishnan80:650,McLean80:5639,Blaudeau97:5016} on all atoms. Additionally, the basis set was augmented with 2 sets of s, p, d diffuse functions in the case of \ki, and 3 sets of s, p, d diffuse functions in the case of \cli. The active space comprises the 2s and 2p orbitals of K/Cl with occupancy fixed to 6 and all virtual orbitals with occupancy fixed to 1. The remaining doubly occupied orbitals were frozen in the calculation. \citep{mosnier16:061401}

>>>>

For the present experiment we used the newly operational microjet setup that was specifically designed for the HAXPES station of the GALAXIES beamline \citep{ceolin13:188,rueff15:175}. A differentially-pumped tube in which the microjet head is inserted, is mounted on a 3-axes manipulator in front of the spectrometer lens. Two holes of 2\,mm diameter allow the photons to go in and out. At the end of the tube and in front of the lens, a 500\,$\mu$m diameter hole skimmer allows the electrons created at the interaction point to go in the direction of the spectrometer. The microjet head is mostly composed of a 30\,$\mu$m diameter vertical glass capillary facing a temperature-controlled catcher in CuBe having a 300\,$\mu$m hole, and a camera. Piezo motors allow their precise alignment relative to each other and to the photon beam. The catcher is placed at a distance of about 5\,mm from the capillary and is permanently pumped in order to extract the liquid. For the present experiment, a 0.5M KCl aqueous solution is injected in the capillary by a high performance liquid chromatography (HPLC) pump with a constant flux of 1.6\,ml/min. The alignment of the setup is performed on a salted aqueous solution by measuring the O1s XPS peak intensity and by optimizing the liquid phase vs gas phase ratio. The pressure in the main chamber is kept below the 10$^{-5}$\,mbar range whereas it is kept at about 10$^{-4}$\,mbar in the differentially-pumped tube when the HPLC pump is ON. Our equipment is an updated version of the equipment used in the reference \cite{faubel88:269}. The aqueous potassium chloride solution was prepared by mixing >99\% KCl salt with deionized water. Filtering and degazing procedures were systematically performed before injecting the solution. The spectrometer resolution of about 0.6\,eV was achieved with the 500\,eV pass energy and 0.5\,mm slits. The photon energy resolution achieved at 2.8\,keV and 3.6\,keV was about 0.3\,eV and 0.4\,eV, respectively. The experimental 2D maps representing the dependence of the electron kinetic energy on the photon energy of KCl at the K-edges of \ki~and \cli, are shown in Figs.\ \ref{fg:2dmap_k} and \ref{fg:2dmap_cl}, respectively.

@@

The theoretical X-ray absorption spectra were computed for the hexa-coordinated clusters of both ions, \ki(H$_2$O)$_6$ and \cli(H$_2$O)$_6$, which can be considered as representatives of the complete first solvation shell of the two ions \citep{Ohtaki93:1157,soper06:180,ma14:1006}. The two structures were optimized at the DFT level of theory using the B3LYP functional and the 6-311++G(2d,2p) basis set \citep{Krishnan80:650,Blaudeau97:5016}. The geometry optimization was performed with the Gaussian 09 package \citep{g09}. In order to obtain more realistic structures corresponding to the bulk solution, we carried out constrained geometry optimization starting with the equilibrium gas-phase geometries \citep{lee99:3995}{\color{red}(REF for Cl-)} belonging to the D$_3$ point group and increasing the angle $\theta$ between the K-O/Cl-O bond and the $C_3$ axis to 55$^{\circ}$ in the case of \ki~and 50$^{\circ}$ in the case of \cli. This angle was chosen such that the O-K-O and O-Cl-O angles are around the maxima of the angular distributions obtained from quantum mechanics / molecular mechanics dynamical simulations in Ref.\ \citep{ma14:1006}. Moreover, we fixed the K-O and Cl-O distances to 2.840~\AA~and 3.140~\AA, respectively, such that they correspond to the distances from other theoretical and experimental works \citep{ge13:13169,gora00:7,Ohtaki93:1157,soper06:180,ma14:1006}. The hexa-coordinated structures are presented in Fig.\ \ref{fg:xas_kcl}. %In both cases, they belong D$_3$ point group \citep{rao08:12944,ge13:13169}.

The energies and transition moments of the core excited states of the microsolvated clusters were computed with the algebraic diagrammatic construction method for the polarization propagator \citep{sch82:2395} within the core-valence separation approximation \citep{bar85:867,ced80:206,ced81:1038} (CVS-ADC(2)x) as implemented in the Q-Chem package \citep{Wenzel14:1900,Wenzel14:4583,Wormit14:774,QChem2015}. In the case of \cli~the 6-311++G(3df,3pd) basis set \citep{Krishnan80:650,McLean80:5639} (excluding the f functions) was used for all atoms, whereas in the case of \ki~we used the 6-311+G(2d,p) basis set \citep{Krishnan80:650,Blaudeau97:5016} on all atoms, and two additional sets of s, p and d diffuse functions were added on K. {\color{red}The use of a smaller basis set in the case of K is due to the higher number of virtual orbitals compared to the case of Cl, and therefore, prohibitively high cost of the CVS-ADC(2)x computation.}
In our calculations the core space comprises the 1s orbital of K$^{+}$ or \cli, whereas the remaining occupied orbitals are included in the valence space. For the calculations of the XAS spectra we used the C$_2$ point group in the case of K$^{+}$(H$_2$O)$_6$ and \cli(H$_2$O)$_6$. To account for the lifetime broadening due to the Auger decay of the core excited states, we convolved the theoretical spectra with a Lorentzian function of FWHM 0.74\,eV and 0.62\,eV in the case of \ki~and \cli, respectively \citep{Krause79:329}. We analyzed the core excited states by expanding the singly occupied natural orbitals (SONOs) $\psi_{i}$ of the microsolvated clusters in the basis of SONOs of the bare K$^{+}$ or \cli~ion $\chi_{nl}$
%
\begin{equation}\label{eq:sono_proj}
\psi_{i} = \sum_{nl} a^{i}_{nl} \chi_{nl}
\end{equation}
%
where $n$ and $l$ stand for the principal and orbital quantum numbers as described in Ref.\ \citep{miteva16:16671}. The expansion coefficients $a^{i}_{nl}$ show the degree of delocalization of the excited electron and the mixing of the core excited states in the crystal field created by the surrounding water molecules (see Fig.\ \ref{fg:xas_kcl}).

The final states following KLL resonant Auger decay of K$^{+}$(H$_2$O)$_6$ and \cli(H$_2$O)$_6$ were computed at the Configuration Interaction Singles (CIS) level using the Graphical Unitary Group Approach (GUGA) as implemented in the GAMESS-US package \citep{GUGA_PhysScr_21,GUGA_JCP_70,GUS}. In order to account for the relaxation effects upon core ionization, we used a restricted open-shell Hartree-Fock reference wave function with a hole in the 2s orbital of both \ki~and \cli.  We used the 6-311++G(2d,2p) basis set \citep{Krishnan80:650,McLean80:5639,Blaudeau97:5016} on all atoms. Additionally, the basis set was augmented with 2 sets of s, p, d diffuse functions in the case of \ki, and 3 sets of s, p, d diffuse functions in the case of \cli. {\color{red}The larger basis set employed in the case of Cl was necessary in order to ensure the convergence of the excited states.}
The active space comprises the 2s and 2p orbitals of K/Cl with occupancy fixed to 6 and all virtual orbitals with occupancy fixed to 1. The remaining doubly occupied orbitals were frozen in the calculation. \citep{mosnier16:061401}
